The Design, Semantics, and Implementation of the Synchronous Language \CEU

### 2016-03-26 paper.tex
...

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C via function annotations and enforced 
finalization for external resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
% TODO: why? stack, fin, principalmente
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing that the peculiarities in the semantics of \CEU do not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU goes one step further and enforces the use of \code{finalize} for system 
calls that deal with pointers representing resources:
%
\begin{itemize}
\item If a system call \textbf{receives} a pointer from \CEU, the pointer 
represents a \textbf{local} resource that requires finalization.
\item If a system call \textbf{returns} a pointer to \CEU, the pointer 
represents an \textbf{external} resource that requires finalization.
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} 
returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await 1s} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile if the programmer forgets to use the
\code{finalize} construct.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{(a) Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await 1s;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{(b) External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

@@

@@

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
TODO: we could detect that trails do not share memory and execute in multiple 
threads

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event.
For this reason, \CEU{} does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

@@

\CEU has a strong influence from Esterel and embraces the disciplined 
synchronous-reactive model with support for lexical composition of lines of 
execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel.
%
In particular, Esterel has a notion of time similar to digital circuits in 
which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provide a limited form of coroutines supporting reactive statements.

>>>>

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward single-threaded implementation targeting resource-constrained 
systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C via function annotations and enforced 
finalization for external resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
% TODO: why? stack, fin, principalmente
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing that the peculiarities in the semantics of \CEU do not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU goes one step further and enforces the use of \code{finalize} for system 
calls that deal with pointers representing resources:
%
\begin{itemize}
\item If a system call \textbf{receives} a pointer from \CEU, the pointer 
represents a \textbf{local} resource that requires finalization.
\item If a system call \textbf{returns} a pointer to \CEU, the pointer 
represents an \textbf{external} resource that requires finalization.
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} 
returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile if the programmer forgets to use the
\code{finalize} construct.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{(a) Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{(b) External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

@@

TODO add a remark about tasks/async/thread in "Discussion"

@@

On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.

TODO: we could detect that trails do not share memory and execute in multiple 
threads

%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

\CEU favors a fine-grained use of trails, being common to use trails that await 
a single event.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm,wsn.ocram}: memory 
for blocks in sequence are packed in a \code{struct}, while blocks in parallel, 
in a \code{union}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at a given time.
A given position in the memory may hold different data (with variable sizes) 
during runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with corresponding 
memory layout.
%
Each variable is assigned a unique $id$ (e.g. \code{a\_1}) so that variables 
with the same name can be distinguished.
%
The \code{do-end} blocks in sequence are packed in a \code{union}, given that 
their variables cannot be in scope at the same time, e.g., \code{MEM.a\_1} and 
\code{MEM.b\_2} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags related to the 
parallel composition, which also reside in reusable slots in the static memory.

@@

\subsection{Single-Threaded Dispatching}

The implementation of \CEU dispatches active trails sequentially in a single 
thread, taking no advantage of multi-core CPUs.
%
This decision comes not only from the fact that \CEU targets constrained 
embedded systems with a single CPU, but also because \CEU imposes deterministic 
execution for intra-reaction statements.

Nonetheless, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
analysis of \CEU infers precisely trails that are concurrent yet do not share 
resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with a multi-threaded implementation in multi-core 
CPUs execute slower than the single-threaded implementation in the same CPUs.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising~\cite{esterel.multicore.1,esterel.multicore.2}.
We believe that
    the continuous fork-and-rejoin overhead due to small reactions
as well as
    the excessive locality of data due to stackless threads sharing contiguous static memory
seem to prevent any speed-up gains.

If we consider data-dominated applications, multi-core implementations can 
offer considerable speed-up gains.
However, data-intensive computations do not typically require a disciplined 
step-wise execution and can actually execute in asynchronous calls.
Esterel provides a \code{task} primitive for this 
purpose~\cite{esterel.primer}, and \CEU provides an equivalent 
\code{async/thread} primitive (which are out of the scope of this paper).

@@

\CEU has a strong influence from Esterel, embracing its disciplined 
synchronous-reactive model with support for lexical composition of lines of 
execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel.
%
In particular, Esterel has a notion of time similar to digital circuits in 
which multiple signals can be active at a clock tick.
%In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provide a limited form of coroutines supporting reactive statements.
